---
title: R companion to applied reg, ch3 - Exploring and Transforming data
Date: 2025/10/26
format: html
---
* Plotting needed at 3 stages:
  - EDA (term courtesy of Tukey in 1977): To get a general idea of the outliers, odd and interesting features.
  - checking model fit
  - presenting model to others

```{r}
#| message: false
library("car")
Prestige[sample(nrow(Prestige), 10),]
```
* Histogram: default bins vs *Freedman & Diaconis(1981)* method

```{r}
hist(Prestige$income)
hist(Prestige$income, breaks = "FD" )
box()
```
* printing all parameters of a function

```{r}
args("plot.default")
```
* `stem.leaf()` function in `aplpack` package for pretty-printed histograms

```{r}
library("aplpack")
sl <- stem.leaf(Prestige$income)
stem(Prestige$income)
```
* density estimation with the `density()` function

density estimate is overlayed on a *density histogram*. In a density histogram, 
bin size is kept variable so that the area of a column is the probability of the bin
We plot this using `freq=F`
```{r}
hist(Prestige$income, freq=F, breaks="FD")
rug(Prestige$income)
lines(density(Prestige$income, from=0), lwd=2, lty="dashed")
legend("topright", legend="density", lty = "dashed", lwd = 2, inset=0.02)
box()
```


