---
title: Advanced R
format: 
  html:
    css: styles.css
    number-sections: true
  pdf:
    include-in-header:
      text: |
        \input{preamble.tex}
---


```{r}
#| include: false
knitr::opts_chunk$set(
  results = "markup",
  collapse = TRUE,
  comment = ""
)
```
# Memory Management

```{r}
#| include: false
## library to inspect objects
library(lobstr)
library(tibble)
```

```{r}

a <- runif(1e6)
as.integer(obj_size(a))
as.integer(obj_size(list(a, a)))
```


(@) reassignment [does not create]{.warning} new objects.

```{r}
x <- c(1,2,3)
y <- x
obj_addr(x)
obj_addr(y)
ref(x)
ref(y)
```
- Can use either `obj_addr` or `ref`

(@) strict about syntactic names: only [characters]{.key}, [underscore]{.key}, [digits]{.key} and [dots]{.key} are allowed.
Can [only]{.warning} start with [digits]{.warning} or [dot but not followed by a number]{.warning}. 
In order to use [non-syntactic]{.warning} identifiers, use [backticks.]{.warning}
    
```{r}
`_abc` <- 1
`_abc`
`10` <- 15
`10`
```
(@) use `make.names` to convert to valid values. 
    
(@) use `check.names = F` option with `read.csv()` since we don't want R to convert column names internally using `make.names()`

(@) [copy-on-modify]{.key} objects are copied only when they are modified

    * Once duplicate object was created for y, it is the [single pointer]{.warning} to the new object. So, subsequent
changes [do not create]{.warning} any further objects. This is an example of [modify-in-place.]{.key}
```{r}
x <- c(1,2,3)
cat(tracemem(x), "\n")
y <- x # not here
y[[2]] <- 22
y[[3]] <- 24
untracemem(y)
```

(@) copy-on-modify triggered only when there are more than one name bound to an object. So, the following does not trigger any copies.

```{r}
x <- c(1, 2, 3)
cat(tracemem(x), "\n")
x[[3]] <- 4
untracemem(x)
```
(@) function arguments follow copy-on-modify rule on the passed parameter too.
    
    * if the passed parameter is returned back without modification, this would be the same pointer that was passed.

(@) vector objects are created with values in place whereas [lists]{.warning} have [pointers]{.warning} to 
values created separately.

```{r}
vec <- c(1,2,3)
l <- list(1,2,3)

ref(vec)
ref(l)
```
(@) [copy-on-modify for lists]{.key} When lists are modified, they are [shallow copied]{.warning}.
This means that the copied list points to the old objects that were not changed. Only change is to the object that was.

```{r}
l1 <- list(1,2,3)
l2 <- l1
l2[[2]] <- 22
ref(l1, l2)
```
(@) [copy-on-modify for data frames]{.key} Dataframes has one pointer for each column 
    
    * [if column is modified,]{.key} copy-on-modify would mean a new dataframe will created with new references for each column. 
Unchanged columns will be shallow-copied i.e. reference to the unchanged column will be to the old memory of the column.
    * [if row modified,]{.key} Deep copy will occur. The new dataframe will have reference to two brand new memory addresses.

```{r}
df <- data.frame(a = 1:5, b = 11:15)
df1 <- df
df1[,2] <- df1[,2] * 2
ref(df, df1)
df2 <- df
df2[2,] <- df2[2,1] * 3
ref(df, df2)
```
(@) character vectors do not store values. They point to a [global pool of strings]{.warning}

```{r}
x <- c("a", "a", "abc", "ac")

ref(x, character = TRUE)
```

(@) [No use declaring an integer vector, since if any member is inadvertently allocated a double value,
it triggers a copy action whereby a double vector is created upcasting each value in the original 
]{.warning}
```{r}
x <- c(1L, 2L, 3L)
tracemem(x)
x[[3]] <- 4
untracemem(x)
```
(@) [Address extensions are not reported by `tracemem`]{.warning}. Note below that `x` was copied when 
we attempted to extend its size to 3. This is [true for both vectors and lists]{.warning}.
```{r}
x <- list(1, 2)
obj_addr(x)
tracemem(x)
x[[3]] <- 3
obj_addr(x)
untracemem(x)

x <- c(1, 2)
obj_addr(x)
tracemem(x)
x[[3]] <- 3
obj_addr(x)
untracemem(x)
```

[Q. What happens when a list is allocated to itself as a member?]{.question}
shallow-copy of x pointing to the old copy of x. [Note: Ignore the extra copies internally created by
`ref` triggering the two extra `tracemem` messages.]

```{r}
x <- list(1,2)
ref(x)
tracemem(x)
x[[2]] <- x
ref(x)
```
(@) Repeated objects are counted only once

```{r}
x <- runif(1e6)
obj_size(x)
obj_size(list(x,x,x))
```
(@) [copy-in-place]{.key} when object is bound to only one name. Also, [environments]{.warning} are always modified in place.

[Q. What happens when `df[[1]] <- df[[1]] + 0.5` is called]{.question}
```{r}
df <- data.frame(x = 1:10, y = 11:20, z = 21:30)

cat(tracemem(df), "\n")
df[[1]] <- df[[1]] + 0.5
untracemem(df)

```

R's [replacement function machinery]{.key} kicks in. The statement is replaced by 
[df <- `[[<-.data.frame`(df*, i, value)]{.warning}.

R creates 2 copies of the object:

- [original df ──copy──► temporary df*]{.warning} before entering the replacment function. This is special behaviour of the replacement machinery and not how it works for ordinary functions.
- [`[[<-.data.frame` is like an ordinary function whose argument x creates another reference beside the df* retained by the replacement machinery]{.warning}. So, when it is modified another 
copy is created.

[Q. If there are many columns are they need to be modified in a loop, what's the recommended pattern?]{.question}

Process the modification in a batch using `vapply` or `Map` and apply at a single shot using `df[]`. This will make the replacement machinery trigger only once.

```{r}
df <- data.frame(matrix(runif(5 * 10), ncol = 5))
medians <- vapply(df, median, numeric(1))

cat(tracemem(df), "\n")

df[] <- Map(\(x, m_i) x + m_i, df, medians)
# df[] <- vapply(df, \(x) x + 0.5, numeric(1))
untracemem(df)
```

# Vectors

(@) [Vector hierarchy]{.key}

```{mermaid}
flowchart BT
Date --> Integer
POSIXct --> Double
difftime --> Double
Logical --> Atomic
Integer --> Numeric
Double --> Numeric
Numeric --> Atomic
Character --> Atomic
Atomic --> Vector
data.frame --> List
tibble --> List
List --> Vector
NULL

Logical["**Logical**"]
Integer["**Integer**"]
Double["**Double**"]
Character["**Character**"]
List["**List**"]


Note["bold: Nodes that typeof() returns and those for which is.* function is safe to use"]
style Note fill:#fff7ed,stroke:#ea580c,color:#9a3412
```

(@) Vector concatenation [flattens]{.warning} into a single vector.


```{r}
c(c(1,2,4), c(10, 11, 12))
```

(@) `typeof()` for getting the type of the vector. [Do not use]{.warning} `mode()` or `storage.mode()` as they are there for old code compatibility purposes.

```{r}
typeof(c(1L))
typeof(c(1))
typeof(c(TRUE))
typeof(c("Hello"))
```
(@) Only use the [lowest level type checkers]{.warning}. 

```{r}
is.integer(1L)
is.integer(1)
is.double(1)
is.logical(1)
is.logical(T)
is.character("1")
is.character(1)
```

(@) Every operation with `NA` leads to na [other than]{.warning} `NA^0`, `NA | TRUE` , `NA & FALSE`

(@) Test `NULL` and `NA` with `is.null()` and `is.na()`.

```{r}
is.null(names(1:10))
is.na(NA)
```
(@) Coercion rule: [Character > Double > Integer > Logical]{.key}.


```{r}
typeof(c(1L, TRUE))
typeof(c(1, 1L, TRUE))
typeof(c("1", 1, 1L, TRUE))
```

[Q. Use of `is.atomic()`, `is.numeric()` and `is.vector()`?]{.question}
- `is.atomic()`: Is it the object an atomic vector OR is it NULL?
- `is.numeric()`: Numeric vector (integer or double) AND NOT ANY(`factor`, `Date`, `difftime`, `POSIXt`)
- `is.vector()`: Vector AND no attributes other than names


(@) [Attributes]{.key}: Attributes are `Pairlist`. Get with `attr()` and `attributes()`. Set with `structure()`.
```{r}
x <- c(1,2,3)
is.vector(x)
x <- structure(
  x,
  a = "first",
  b = 2L
)
attributes(x)
is.vector(x)

```
(@) The `dim` attribute turns an atomic vector into a matrix. So, it is set using `structure()` the atomic vector can be converted to 
 matrix within using `matrix()` function.

```{r}
m <- structure(
  1:10,
  dim = c(5,2)
)

is.matrix(1:10)
is.matrix(m)
m
```
[Q. Common methods of atomic vectors, matrices, and array]{.question}
- *Vectors*: `c()`, `names()`, `length()`, `is.null(dim(x))`
- *Matrices*: `is.matrix()`, `colnames()`, `rownames()`, `nrow()`, `ncol()`, `rbind()`, `cbind()`, `t()`
- *Arrays*: `dimnames()`, `dim()`, `is.array()`

:::{.definition}
**S3 Atomic vectors** Vectors with `class` attribute. Ex. `factor`,`Date`, `POSIXct`, `difftime`. 
Some of these S3 vectors have other attributes as well.
:::




(@) `factor` are built over `integer` types, has `class` and `levels` attributes. [Non-existent
values]{.warning} can be part of the levels as well.

```{r}
x <- c("b", "a", "a", "a", "b")
table(x)
x_f <- factor(x, levels = c("a", "b", "c"))
table(x_f)
attributes(x_f)
```
(@) `Date`, `POSIXct` and `difftime` built over `double`. First two are no. of days and no. of seconds
since *1970-01-01* and the latter uses `unit` to store durations.
```{r}
today <- Sys.Date()
today
str(today)
dt <- as.Date("1970-02-01")
unclass(dt)

tm <- as.POSIXct("1970-01-02 12:00", tz = "UTC")
attributes(tm) <- NULL
tm

wk_1 <- as.difftime(20, units = "weeks")
str(wk_1)
```
- use `attributes(x) <- NULL` to get their undelying values.

[Q. How does `level` construction and array construction happen in factor?]{.question}

```{r}
f3 <- factor(letters, level = rev(letters))
f3
as.integer(f3)
```
 1. Numeric [level]{.warning} construction happens first. The numeric levels are from 1-26.
 2. The numeric levels are assigned to the [text labels]{.warning}. So, 1 is mapped to 'z', 2 to 'y' and so on. This is because 'letters' has been reversed.
 3. At this point the main vector is constructed. Since there exists a level, the first value 'a' is looked up against the labels of the level. It finds the corresponding
 level as '26' so it stores 26 and so on.

[Q. What happens when labels are reassigned using `levels()`]{.question}

```{r}
f <- factor(c("a", "b", "b", "c", "a", "b"))
f
levels(f) <- c("b", "c", "a")
f
```
 1. `levels` are constructed in the first `factor()` call along with labels. So, 1 $\rightarrow$ "a", 2 $\rightarrow$ "b"
 2. Main vector stores integer values (1,2,2,3,1,2)
 3. In the `levels(f)` call, levels are [re-leveled]{.warning}. So, 1 $\rightarrow$ "b", etc. 
 4. Finally, when `f` is printed, it prints the labels of the main vector based on the new labelling.

:::{.insight}
Factor levels often need to be rearranged so that the reference can be a particular category when dummy variables are defined during a regression exercise.
The statement `factor(x,levels=` does this correctly while the statement `levels(x) <-` means "rename the old labels into these new labels", which is wrong.
The **best** option is to make the intention clear by using `relevel(df$x, ref = "c")`
:::

```{r}
df <- data.frame(y = 1:10, x = c("a", "b", "a", "c", "a", "a", "b", "c", "c", "b"))

# CORRECT
# df$x <- factor(df$x, levels = c("c", "b", "a"))

# WRONG
# df$x <- factor(df$x)
# levels(df$x) <- c("c", "b", "a")

# BEST
df$x <- factor(df$x)
df$x <- relevel(df$x, ref = "c")
# str(df)
# df
model.matrix(~x, data = df)
```

[Q. How to convert a list to an atomic vector?]{.question} `unlist(lst)`. `unlist` strips **all** 
attributes and retains only the underlying value. The common type is the broadest type of all. 

```{r}
l <- list(a = 1:5, b = c("a", "b", "c"))
str(l)
typeof(l)
ref(l)
l <- unlist(l)
str(l)
typeof(l)
ref(l)
```

[Q. `typeof` on list returns?]{.question} `list`

[Q. Difference between `c(as.Date("1970-01-02"), as.POSIXct("1970-01-01 02:00", tz = "UTC"))` and `unlist(list(<a date>,<a posixct>))`?]{.question}

- `c`:
  1. considers what the type of the inputs are. In this case, they are `Date` and `POSIXct` respectively.
  2. Since, the first argument is Date, it retains the Date part of both i.e. converts the later to date.
  3. Retains the internal integer representation of date.

```{r}
d <- as.Date("1970-01-02")
dt <- as.POSIXct("1970-01-01 01:00", tz = "UTC")
unclass(c(d, dt))
# In this case d is converted to posixct and internally the double representation 
# stores the number of seconds.
unclass(c(dt, d))
```
- `unlist()`
  1. Strips all special information stored with the date/datetime objects.
  2. No date conversations as before. Just retains the [original]{.warning} double/integer values.

```{r}
unclass(unlist(list(d, dt)))
unclass(unlist(list(dt, d)))
```


(@) use `as_tibble()` to convert from `data.frame` to `tibble`

## Key features of `data.frame`

(1) [Both matrix-like and list-like]{.warning}: `rownames()` , `colnames()`, `nrow()`, `ncol()` work but so does `length()` which returns [no. of columns]{.warning}.
(2) `is.data.frame()` returns true for [both]{.warning} dataframe and tibble. But `is.tibble()` returns true [for only]{.{arning} tibbles.

## `tibble` > `data.frame`

(1) [No auto-conversion to factor]{.key} `stringsAsFactors = FALSE` [not needed]{.warning} to prevent string columns from being automatically converted to factors.

(2) [non-syntactic names]{.key} are not replace in tibbles.

(3) [recycling]{.key} only works with columns of size 1 in tibbles.

```{r}
# Data frame recycled if the largest one is 
# a multiple of all the other smaller ones. So, the below works fine.
df <- data.frame(x = 1:6, y = 1:2)
df

# This will throw an error
# tbl <- tibble(x = 1:10, y = 1:2)
```
(5) [columns during construction]{.key} Tibble allows reference to columns defined during construction.

```{r}
df <- tibble(
x = 1:3,
  y = x * 3
)
df
attributes(df)
```
(6) [auto conversion of single column selection]{.key} Single column selected `[` is [not converted automatically to vector]{.warning} in tibbles.

```{r}
df <- data.frame(
x = 1:3,
y = 11:13
)
# Autoconversion of single columns to vectors. This does not happen with tibbles.
df[,"y"]
df <- as_tibble(df)
df[,"y"] # Still a tibble
```
:::{.insight}
Use the list-style column access method `df[["y"]]` to directly get the **vector version** of a column. This makes intention clear. It works for **both `data.frame` and `tibble`**.
Equally, **always** use **drop = FALSE** with data.frame single columns access `df[,"y", drop = F]`
:::

```{r}
df <- tibble(
x = 1:3,
y = 11:13
)
df[["y"]]
```

(7) [simpler list-as-column].{key} Including list as a column requires less code and is printed better

```{r}
df <- tibble(
x = 1:3,
y = list(1:2, 1:3, 1:4)
)
df
```
[Q. How do we include `list`, `matrix`, `array` in `data.frame`]{.question} 

Either [define separately]{.warning} or [use `I`]{.warning}. `I` means "leave it as-it-is."

```{r}
# This won't work
# df <- data.frame(x = 1:3, y = list(1:3, 1:4, 1:5))

# OPTION 1:
df <- data.frame(x = 1:3)
df$y = list(1:3, 1:4, 1:5)
df

# OPTION 2:
df <- data.frame(x = 1:3, y = I(list(1:3, 1:4, 1:5)))
```
(8) [Row names not supported]{.key} Tibbles do not support row names. We can use `rownames` attribute to convert to a column.


```{r}
df <- data.frame(x = 1:3, y = 21:23, row.names = c("first", "second", "third"))
tb <- as_tibble(df, rownames = "names")
tb
```

[Q. What happens when we transpose a dataframe?]{.question} data.frame is converted into a matrix.

[Q. Whats the difference between `as.matrix()` and `data.matrix()`]{.question}

`as.matrix()` uses the Char > Double > int > logical hierarchy. `data.matrix()` converts the values into [numeric values]{.warning}.

`data.matrix()` rules:

  1. *Characters*: convert into factors and return their internal values.
  2. *Factors*: Return internal integer values. 


```{r}
df <- data.frame(
  a = c("a", "b"),
  b = c(TRUE, FALSE),
  c = c(1L, 0L),
  d = c(1.5, 2),
  e = factor(c("f1", "f2"))
)

am <- as.matrix(df)
dm <- data.matrix(df)
am
typeof(am)
dm
typeof(dm)
```




