---
title: Advanced R
format: 
  html:
    css: styles.css
  pdf:
    include-in-header:
      text: |
        \input{preamble.tex}
---


```{r}
#| include: false
knitr::opts_chunk$set(
  results = "markup",
  collapse = TRUE,
  comment = ""
)
```
#### Imports
```{r}
## library to inspect objects
library(lobstr)
library(tibble)
```

```{r}

a <- runif(1e6)
as.numeric(obj_size(a))
as.numeric(obj_size(list(a, a)))
```


(@) reassignment [does not create]{.warning} new objects.

```{r}
x <- c(1,2,3)
y <- x
obj_addr(x)
obj_addr(y)
ref(x)
ref(y)
```
- Can use either `obj_addr` or `ref`

(@) strict about syntactic names: only [characters]{.key}, [underscore]{.key}, [digits]{.key} and [dots]{.key} are allowed.
Can [only]{.warning} start with [digits]{.warning} or [dot but not followed by a number]{.warning}. 
In order to use [non-syntactic]{.warning} identifiers, use [backticks.]{.warning}
    
```{r}
`_abc` <- 1
`_abc`
`10` <- 15
`10`
```
(@) use `make.names` to convert to valid values. 
    
(@) use `check.names = F` option with `read.csv()` since we don't want R to convert column names internally using `make.names()`

(@) [copy-on-modify]{.key} objects are copied only when they are modified

    * Once duplicate object was created for y, it is the [single pointer]{.warning} to the new object. So, subsequent
changes [do not create]{.warning} any further objects. This is an example of [modify-in-place.]{.key}
```{r}
x <- c(1,2,3)
cat(tracemem(x), "\n")
y <- x # not here
y[[2]] <- 22
y[[3]] <- 24
untracemem(y)
```

(@) copy-on-modify triggered only when there are more than one name bound to an object. So, the following does not trigger any copies.

```{r}
x <- c(1, 2, 3)
cat(tracemem(x), "\n")
x[[3]] <- 4
untracemem(x)
```
(@) function arguments follow copy-on-modify rule on the passed parameter too.
    
    * if the passed parameter is returned back without modification, this would be the same pointer that was passed.

(@) vector objects are created with values in place whereas [lists]{.warning} have [pointers]{.warning} to 
values created separately.

```{r}
vec <- c(1,2,3)
l <- list(1,2,3)

ref(vec)
ref(l)
```
(@) [copy-on-modify for lists]{.key} When lists are modified, they are [shallow copied]{.warning}.
This means that the copied list points to the old objects that were not changed. Only change is to the object that was.

```{r}
l1 <- list(1,2,3)
l2 <- l1
l2[[2]] <- 22
ref(l1, l2)
```
(@) [copy-on-modify for data frames]{.key} Dataframes has one pointer for each column 
    
    * [if column is modified,]{.key} copy-on-modify would mean a new dataframe will created with new references for each column. 
Unchanged columns will be shallow-copied i.e. reference to the unchanged column will be to the old memory of the column.
    * [if row modified,]{.key} Deep copy will occur. The new dataframe will have reference to two brand new memory addresses.

```{r}
df <- data.frame(a = 1:5, b = 11:15)
df1 <- df
df1[,2] <- df1[,2] * 2
ref(df, df1)
df2 <- df
df2[2,] <- df2[2,1] * 3
ref(df, df2)
```
(@) character vectors do not store values. They point to a [global pool of strings]{.warning}

```{r}
x <- c("a", "a", "abc", "ac")

ref(x, character = TRUE)
```

(@) [No use declaring an integer vector, since if any member is inadvertently allocated a double value,
it triggers a copy action whereby a double vector is created upcasting each value in the original 
]{.warning}
```{r}
x <- c(1L, 2L, 3L)
tracemem(x)
x[[3]] <- 4
untracemem(x)
```
(@) [Address extensions are not reported by `tracemem`]{.warning}. Note below that `x` was copied when 
we attempted to extend its size to 3. This is [true for both vectors and lists]{.warning}.
```{r}
x <- list(1, 2)
obj_addr(x)
tracemem(x)
x[[3]] <- 3
obj_addr(x)
untracemem(x)

x <- c(1, 2)
obj_addr(x)
tracemem(x)
x[[3]] <- 3
obj_addr(x)
untracemem(x)
```

[Q. What happens when a list is allocated to itself as a member?]{.question}
shallow-copy of x pointing to the old copy of x. [Note: Ignore the extra copies internally created by
`ref` triggering the two extra `tracemem` messages.]

```{r}
x <- list(1,2)
ref(x)
tracemem(x)
x[[2]] <- x
ref(x)
```
(@) Repeated objects are counted only once

```{r}
x <- runif(1e6)
obj_size(x)
obj_size(list(x,x,x))
```
(@) [copy-in-place]{.key} when object is bound to only one name. Also, [environments]{.warning} are always modified in place.

(@) [Q. What happens when `df[[1]] <- df[[1]] + 0.5` is called]{.question}
```{r}
df <- data.frame(x = 1:10, y = 11:20, z = 21:30)

cat(tracemem(df), "\n")
df[[1]] <- df[[1]] + 0.5
untracemem(df)

```

R's [replacement function machinery]{.key} kicks in. The statement is replaced by 
[df <- `[[<-.data.frame`(df*, i, value)]{.warning}.

R creates 2 copies of the object:

- [original df ──copy──► temporary df*]{.warning} before entering the replacment function. This is special behaviour of the replacement machinery and not how it works for ordinary functions.
- [`[[<-.data.frame` is like an ordinary function whose argument x creates another reference beside the df* retained by the replacement machinery]{.warning}. So, when it is modified another 
copy is created.

(@) [Q. If there are many columns are they need to be modified in a loop, what's the recommended pattern?]{.question}

Process the modification in a batch using `vapply` or `Map` and apply at a single shot using `df[]`. This will make the replacement machinery trigger only once.

```{r}
df <- data.frame(matrix(runif(5 * 10), ncol = 5))
medians <- vapply(df, median, numeric(1))

cat(tracemem(df), "\n")

df[] <- Map(\(x, m_i) x + m_i, df, medians)
# df[] <- vapply(df, \(x) x + 0.5, numeric(1))
untracemem(df)
```

### 2. Vectors

:::{.definition}
**S3 Atomic vectors** Vectors with `class` attribute. Ex. `factor`,`Date`, `POSIXct`, `difftime`. 
Some of these S3 vectors have other attributes as well.
:::

(@) [Vector hierarchy]{.key}

```{mermaid}
flowchart BT
Logical --> Atomic
Integer --> Numeric
Double --> Numeric
Numeric --> Atomic
Character --> Atomic
Atomic --> Vector
List --> Vector
NULL

```

(@) Vector concatenation [flattens]{.warning} into a single vector.


```{r}
c(c(1,2,4), c(10, 11, 12))
```

(@) `typeof()` for getting the type of the vector. [Do not use]{.warning} `mode()` or `storage.mode()` as they are there for old code compatibility purposes.

```{r}
typeof(c(1L))
typeof(c(1))
typeof(c(TRUE))
typeof(c("Hello"))
```
(@) Only use the [lowest level type checkers]{.warning}. 

```{r}
is.integer(1L)
is.integer(1)
is.double(1)
is.logical(1)
is.logical(T)
is.character("1")
is.character(1)
```

(@) Every operation with `NA` leads to na [other than]{.warning} `NA^0`, `NA | TRUE` , `NA & FALSE`

(@) Coercion rule: [Character > Double > Integer > Logical]{.key}.


```{r}
typeof(c(1L, TRUE))
typeof(c(1, 1L, TRUE))
typeof(c("1", 1, 1L, TRUE))
```


```{r}
m <- matrix(1:10, ncol = 2)
attrs(m)
is.vector(m)
```


- `factor` are built over `integer` types, has `class` and `levels` attributes. [Non-existent
values]{.warning} can be part of the levels as well.

```{r}
x <- c("b", "a", "a", "a", "b")
table(x)
x_f <- factor(x, levels = c("a", "b", "c"))
table(x_f)
attributes(x_f)
```
- `Date`, `POSIXct` and `difftime` built over `double`. First two are no. of days and no. of seconds
since *1970-01-01* and the latter uses `unit` to store durations.
```{r}
today <- Sys.Date()
today
str(today)
dt <- as.Date("1970-02-01")
unclass(dt)

tm <- as.POSIXct("1970-01-02 12:00", tz = "UTC")
attributes(tm) <- NULL
tm

wk_1 <- as.difftime(20, units = "weeks")
str(wk_1)
```
- use `attributes(x) <- NULL` to get their undelying values.

[Q. How to convert a list to an atomic vector?]{.question} `unlist(lst)`. `unlist` strips **all** 
attributes and retains only the underlying value.

- use `as_tibble()` to convert from `data.frame` to `tibble`

#### `tibble` > `data.frame`

(@) tibble allows reference to columns defined during construction.
```{r}
df <- tibble(
x = 1:3,
  y = x * 3
)
df
attributes(df)
```
(@) single column selected `[` is [not converted automatically to vector]{.warning}.

```{r}
df <- data.frame(
x = 1:3,
y = 11:13
)
df[,"y"]
df <- as_tibble(df)
df[,"y"]
```
(@) Including list as a column requires less code and is printed better

```{r}
df <- tibble(
x = 1:3,
y = list(1:2, 1:3, 1:4)
)
df
```
[Q. How do we get a selected tibble column directly in the form of a vector] `df[["col1"]]`

```{r}
df <- data.frame(
x = 1:3,
y = 11:13
)
df[["y"]]
```

```{r}
a <- numeric()
```
