---
title: R 4 DS, ch1, Plotting using ggplot
format: html
---


```{r}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

* Initial exploration of the penguin tibble
`glimpse()` function like `str()` 

```{r}
head(penguins)
glimpse(penguins)
str(penguins)
```
* Answering questions:
 - relationship between bodymass and flipper length
 - is this relationship different for different types
 - linear or not?


```{r}
penguins |> ggplot(
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) + 
  geom_point() + 
  geom_smooth(method = "lm") 
```
* changing the above to making a single fitting line as opposed to one for each type
  - `color` at a local geom_point level


```{r}

penguinplot <- penguins |> ggplot(
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) + 
  geom_point(
    mapping = aes(color = species, shape = `species`)
  ) + 
  geom_smooth(method = "lm") 

penguinplot
```
* adding labels as a new layer

```{r}
penguinplot <- penguinplot + 
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinsstrap, and Gentoo penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    shape = "Species"
  )
penguinplot
```
* scaling color to safe for color blindness


```{r}
penguinplot +
  scale_color_colorblind()
```
* Some more info about penguins dataset


```{r}
penguins |> nrow()
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm
  )
) + 
  geom_point(
    mapping = aes(
      color = species,
      shape = species
    )
  ) +
  labs(
    title = "Bill depth vs bill length",
    subtitle = "subtitle",
    x = "Bill length(mm)",
    y = "Bill depth(mm)"
  ) +
  scale_color_colorblind()
```

* species versus bill_depth


```{r}

ggplot(
  data = penguins,
  mapping = aes(
    y  = species,
    x = bill_depth_mm
  )
) +
  geom_point(na.rm = T) +
  labs(
    caption = "This is a sample caption" 
  )

```
* concise plot

```{r}
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(na.rm = T) +
  geom_smooth(method = "lm") +
  scale_color_colorblind()
```
* distribution of the categorical variable: `species`


```{r}

penguins |>
  ggplot(aes(x = fct_infreq( species ))) + 
  geom_bar()
```
* creating histogram by first creating 200 bins from a continuous value

```{r}
penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

* creating density plot

```{r}
penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_density() +
  labs(
    x = "Body mass (g)",
    y = "density",
    title = "Density plot for body mass"
  )

```
* creating box plots


```{r}
penguins |>
  ggplot(aes(y = body_mass_g, group = species)) + 
  geom_boxplot()
```
* creating density_plot using `group`. `group` is used don't want color to group.

```{r}
penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_density(aes(color = species), linewidth = 0.7)
```

* adding `color` and `fill` aesthetics
 - Note that we only use `aes()` to **map** if we have a variable. For a fixed value, we set it as a 
    parameter directly

```{r}
penguins |>
  # ggplot(aes(x = body_mass_g, fill = species, color = species)) +
  ggplot(aes(x = body_mass_g)) +
  geom_density(fill = "skyblue", color = "skyblue", 
    alpha = 0.6)
```

* bar plot with 2 categorical variables with percentages for comparison


```{r}
penguins |>
  ggplot(aes(x = island, fill = species)) +
  geom_bar(position = "fill")
```
* using `facet_wrap` for each category
  - use the same field for both shape and color of the points
  - use `facet_wrap()` for other categories

```{r}
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(shape = species, color = species)) +
  facet_wrap(~ island)
```

