---
title: Scribble pad
format: html
---


```{r}
library(tidyverse)
library(palmerpenguins)
library(nycflights13)
library(Lahman)
```
- Distribution of 
    - categorical variables with `geom_bar`
    - numerical variables with `geom_histogram`
    - numerical vs categorical with `geom_boxplot` or `geom_density` + `aes(x = ..., color = species, fill = species), alpha = 0.5`
    - categorical vs categorical  - `geom_bar` with `color = species`


```{r}
glimpse(penguins)
```


```{r}
penguins |>
  group_by(species) |>
  summarize(count = n()) |>
  arrange(desc(count))

penguins |>
  ggplot(aes(x = fct_infreq(species))) +
  geom_bar() +
  theme_minimal()
```


```{r}
hist_bins <- function(binw, ...) {
  p <- penguins |>
    ggplot(aes(x = body_mass_g)) +
      geom_histogram(binwidth = binw, ...) + 
      theme_minimal()
  print(p)
}

bins <- c(20, 2000)
walk(bins, hist_bins, na.rm = T)
```

 Numeric vs categorical


```{r}
penguins |>
  ggplot(aes(x = species, y = body_mass_g)) + 
  geom_boxplot(na.rm = T) + 
  theme_minimal()

penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_density(aes(color = species, fill = species), alpha = 0.5, na.rm = T) + 
  labs(
    x = "Body mass",
    y = "Density",
    title = "Density for each species",
    color = "Species"
  ) +
  theme_minimal()
```

```{r}
penguins |>
  ggplot(aes(x = island)) +
  geom_bar(aes(fill = species), position = "fill") +
  labs(fill = "Especies") + 
  theme_minimal()
```

# Data Transformation

- For the **group by + count()** pattern, use `count()`


```{r}
flights |>
count(origin, dest, sort = T) |>
  arrange(n)
```

## `group_by()`, `summarize()` and `slice()`


```{r}
df <- data.frame(x = 1:10, y = c(1,2,1,3,2,2,3,3,1,2), z = c(10,21, 11, 31, 23, 23, 33, 32, 9, 18))
```
- simple group_by

```{r}
df |>
  group_by(y) |>
  summarize(m = max(z))
```
vs

```{r}
df |>
  group_by(y) |>
  filter(z == max(z))
```
vs

```{r}
df |>
  group_by(y) |>
  arrange(y, desc(z)) |>
  slice_head(n = 2)
```
which is the same as 

```{r}
df |>
  group_by(y) |>
  slice_max(order_by = z, n = 2) |>
  arrange(y)
```


```{r}
flights |>
  distinct(dest) |> nrow()

flights |>
  group_by(dest) |>
  # n = 1 alone means "give us all rows with highest value". So, need to use `with_ties = F`
  slice_max(arr_delay, n = 1) |>
  relocate(dest, arr_delay) |>
  filter(n() > 1) |>
  select(dest) |>
  distinct() |>
  pull(1)
  
```

### How to ungroup? 
In one of the two ways:
- When using `summarize` : `.groups = "drop|drop_last(default)|keep"`
    - `ungroup()` otherwise.

### Insight:
Whenever using grouping, use count i.e. `n()`.


```{r}
batters <-
  Batting |>
    group_by(playerID) |>
    summarize(
      performance = sum(H, na.rm = T) / sum(AB, na.rm = T),
      n = sum(AB, na.rm = T)
    )

batters |>
  filter(n > 100) |>
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = F) +
  theme_minimal()

```

# 5. Data Tidying

## Principles:
    - Every variable is a column; columns should only be variables.
    - Every observation should be a row and vice versa.
    - Every data should be a cell and vice versa.

```{r}
billboard
```
-- The columns have data. So we need `pivot_longer()`


```{r}
billboard |>
  pivot_longer(
    cols = matches("^wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = T
  ) |>
  mutate(
    week = str_replace(week,"wk(\\d)", "\\1")
  )
```

```{r}
df <- data.frame(x = 1:10, x1 = 11:20, x2 = 21:30, x3 = 31:40, x4 = 41:50, y = 1.1:10.1)
df |>
  select(matches("x[1-3]?$"))
df |> rownames()
```

-- `pivot_longer` when colnames have data for 3 columns


```{r}
glimpse(who2)
```


```{r}
who2 |>
  pivot_longer(
    cols = !c(country, year),
    names_to = c("diagnosis", "gender", "age"),
    names_pattern = "([^_]+?)_([^_]+?)_([^_]+)",
    values_to = "count",
    values_drop_na = T
  ) 
```
- When both data and variable are there in the name


```{r}
household
```


```{r}
household |>
  pivot_longer(
    cols = !c(family),
    names_to = c(".value", "count"),
    names_pattern = "([^_]+?)_([^_]+)",
    values_drop_na = T
  )
```

- `pivot_wider()`


```{r}
cms_patient_experience
```


```{r}
cms_patient_experience |>
  pivot_wider(
    id_cols = matches("^org.*"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```
# `ggplot()` - Layers


```{r}
mpg |>
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(
    # aes(linetype = drv, color = drv), 
    aes(group = drv), 
    se = F, method = "loess", formula = "y ~ x"
  ) +
  labs(
    linetype = "Drive", color = "Drive"
  ) +
  theme_minimal()
```

- We can apply same aesthetics to different data


```{r}
mpg |>
  ggplot(aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red"
  ) +
  geom_point(
    data = mpg |> filter(class == "2seater"),
    color = "red",
    size = 3, shape = "circle open"
  )
```

- Density/Histogram are always on the x axis

```{r}
mpg |>
  ggplot(aes(x = hwy)) +
  geom_boxplot()
```
